---
title: "circle_plot"
format: pdf
lang: en-GB
editor: 
  markdown:
      wrap: 82
---

# Circle plot

The motivation for creating a circle plot is to visualize data in relation to
seasons. Seasons are a recurring natural phenomenon that can significantly impact
various aspects of our lives, from weather patterns to human behaviour.
Visualizing data in a circle may offer a novel way to explore seasonal changes.

This approach may be valuable for data sets related to changes over time,
particularly over seasons. Take for instance, Tromsø, it is located above [above
the arctic circle](https://en.wikipedia.org/wiki/Arctic_Circle) and exhibits some
of the most extreme daylight changes - from no sunlight at winter, to only
sunlight at summer. A remarkable change in season which could neatly be displayed in a circular plot. 

For this purpose, I want explore the utility of visualizing a data in a circular
way. We recently had a study that investigated participants when they were sleep
deprived and when they were not. To this end, we gathered data regarding
participants sleep time and wake time, as well as controlling with an Actigraphy.
Although I will not use that data here, the data is well suited to visualize
changes during a season.

With this idea in mind, I have created fake sleep data of X along with the sleep
rise and set times located at Tromsø. The sleep data follow the pattern of the sun
changes throughout the changes along with some additional noise. The goal is to
visualize the sleep time, wake time, and sun rise changes throughout a year (2023)
in a circular way.

## Visualizing

We first load the relevant packages and retrieve the data.

```{r preparation}
#| message: false
library(tidyverse) # libs
load("../data/sleep_data.rdata") # load data
```

To start of, we can visualize the data according to sleep amount throughout the year. 

```{r}
sleep_data |>
  ggplot(aes(x=cumulative_days, y=sleep_amount))+
  geom_point()
```

We can see that the data starts off high (in January) and reduce towards middle (summer) and increase towards the end (winter). To investigate the average throughout the year, we can add a geom_smooth. 

```{r}
sleep_data |>
  ggplot(aes(cumulative_days, sleep_amount))+
  geom_point()+
  geom_smooth()
```

To put our plot in context, we can add the amount of daylight in each day throughout the year.

```{r}
sleep_data |>
  ggplot(aes(cumulative_days, sleep_amount))+
  geom_area(mapping=aes(y=daylight_length_h), fill="yellow", alpha = .5)+
  geom_point()+
  geom_smooth()
```

The yellow area display the number of hours of sunlight that day receives. As we can see, the sunlight varies from nothing (at the start and end of the x axis) to all day (the middle of the x axis).

Alternatively, we can display the amount of dark hours throughout the year. 

```{r}
sleep_data |>
  ggplot(aes(cumulative_days, sleep_amount))+
  geom_area(mapping=aes(y=daylight_length_h), fill="yellow", alpha = .5)+
  geom_ribbon(mapping=aes(ymin=daylight_length_h, ymax=24, 
                          xmin=cumulative_days, xmax=cumulative_days), fill="black", alpha=.5)+
  geom_point()+
  geom_smooth()
```

Now, with this plot, we can more easily discern between darkness and sunlight throughout the year. We will leave some alpha to be able to see the underlying grid. 


```{r}
sleep_data |>
  ggplot(aes(cumulative_days, sleep_amount))+
  geom_ribbon(mapping=aes(ymin=daylight_length_h, ymax=24, 
                          xmin=cumulative_days, xmax=cumulative_days), fill="black", alpha=.5)+
  geom_point()+
  geom_smooth()
```

In this figure we see the amount of darkness each day throughout the year. Combining both may aid in discerning hours light and hours darkness throughout the year. 




  geom_hline(yintercept=24)+
  geom_hline(yintercept=0)+
  coord_cartesian(xlim=c(0,365))+
  scale_x_continuous(breaks = seq(1,365,29))


```{r}
sleep_data |>
  group_by(date_m) |> summarise(m = mean(sleep_amount))
  mutate(month_sum = )
  
```



\* should i have stage wise plot updating? Some text and some plotting? OR use it
more to illustrate visualizing principles (perhaps better than showing "my own
learning curve") Plot

To.do: - Indicate that sleep time is -24 hours.

```{r}
sleep_data |> 
  ggplot(aes(x = count, y = wake_time))+
  geom_area(mapping=aes(x=count, y=daylight_length_h/2.4))+
  geom_point()+
  geom_hline(yintercept=8)+
  geom_vline(xintercept=0)+
  coord_polar(start=pi, direction=-1)+
  geom_point(mapping=aes(x=count, y=24-sleep_time))


```

## Comparison to circilize?????

-   [ ] discuss difference to circilize
