---
title: "circle_plot"
format: pdf
editor: 
  markdown:
      wrap: 82
---
# Circle plot

A circle plot displays information in a circular fashion. 
One of the main idea with using a circle plot is that it can easily visualize the
different seasons. Different segments can be highlighted for each of the different
Time periods can be split in

# 
* although codes exists like circlize // OTHERS???
* GGplot is more adaptible, and flexible compared to circlize.
* it follows Grammer of grahics 


## Running Code

```{r preparation}
library(tidyverse) # libs
load("../data/daylight_length_Tromso_2022_finished.rdata") # load data
```

#* should i have stage wise plot updating? Some text and some plotting? OR use it more to illustrate visualizing principles (perhaps better than showing "my own learning curve")
Plot 


To.do:
- Indicate that sleep time is -24 hours. 

```{r}
sleep_data |> 
  ggplot(aes(x = count, y = wake_time))+
  geom_area(mapping=aes(x=count, y=daylight_length_h/2.4))+
  geom_point()+
  geom_hline(yintercept=8)+
  geom_vline(xintercept=0)+
  coord_polar(start=pi, direction=-1)+
  geom_point(mapping=aes(x=count, y=24-sleep_time))


```







