---
title: "Mutlivariate visualization: continous and continous"
format: pdf
lang: en-GB
editor: 
  markdown:
      wrap: 62
---

```{r start libs}

```


```{r start plot}

```


{{< pagebreak >}}


In these latter exercises, I will visualize the distribution of data. More specifically visualize the distribution of data in relation to each other. In the first (this) exercise, I will visualize two continuous variables, while in the latter I will visualize the distribution of data over a categorical variable. To do this I will use a dataset related to mind wandering where we have two variables that are used to predict mind wandering. These two variables related to behavioural performance during the task and are behavioural variability (BV), a measure of response variation, and approximate entropy (AE) a measure of randomness.

I start by loading the libraries, functions and data. 

```{r}
library(tidyverse)
library(plotly)
read_csv("../data/raw_mwtms_data.csv") -> all_mw_data 
source("../src/transformation.R") # AE calculation
```

I then select the relevant data, in this case, I will only use the data from the "sham" condition. That is the condition in which no manipulation was done to participants. 

```{r }
all_mw_data |> 
  select(-region.y) |>
  filter(stimulation == "sham") -> mw_data

da_pro(mw_data, nback=25) -> trans_mw
trans_mw
```

We can now start plotting the data. There are a only a couple of ways one can visualize the distribution of data over a variable. This could be with a histogram or a density distribution. A histogram summarize the data according to the number of occurances within a certain span. 

```{r histogram}
trans_mw |>
  ggplot(aes(x=zlogapen))+
  geom_histogram()
```

It is perhaps a better idea to use a histogram over more meaningful distances of the variable. For instance, an IQ test, where the ranges are be more meaningful. That is, a 5 point increase in IQ is easier to gauge than a .5 increase in zlogapen. Nevertheless, another way of plotting the distribution could be to use a frequency polygon.

```{r}
trans_mw |>
  ggplot(aes(x=zlogapen)) +
  geom_freqpoly()
```

This plot is essentially the same as the histogram, but mimics the density distributions as shown below. It is, however, important to note that this is not a density distribution, as the density is calculated while this plot shows the raw count of each range of value - the same as the histogram. To visualize the distribution more meaningfully, we could decrease the binwidth or increase the number of bins, such that the range of collected data decrease. 

```{r}
trans_mw |>
  ggplot(aes(x=zlogapen)) +
  geom_freqpoly(bins=100)
```

Indeed, in this plot we get a more neuanced view of how the data is distributed. More specifically, we can see that there are a lot of values close to -2. This may be because of how AE is calculated where we can get a ceiling or flooring effect. Even though this specific value receives a lot of observation, most values are distributed around 0. Indeed, the range of the normal distribution is greater than the sum of this extreme score (see below). Moreover, we can see that certain values receives more counts than others, and this may again stem from a limit of how AE is calculated. 

Indeed, visualizing the raw AE score suggest that we see that most values center around 0 (floor) and the ceiling. This is perhaps most obvious with a density plot (see below). 

```{r}
trans_mw |> 
  ggplot(aes(apen))+ 
  geom_freqpoly(bins=100)
```

```{r}
trans_mw |>
  ggplot(aes(apen))+
  geom_density()
```

To get a better grasp of the distribution over values can be to use a density visualization. Visualizing using a density function will smooth the distribution such that it is easier to see the estimated distribution. (Note: this is visualized with the z scored and log transformed AE score, which more accurately follow a normal distribution - compare the above with the below)

```{r density }
trans_mw |>
  ggplot(aes(x=zlogapen))+
  geom_density()
```

A density plot gives us an estimated probability that a data point will be found within that value. With this plot, it is more evident that the data is distributed according to a normal distribution - at least following the relevant transformations (z-score of the -log(log(2)-apen)). 

By z-scoring AE and BV we can meaningfully plot both distributions in the same plot.  

```{r}
trans_mw |>
  pivot_longer(c(zlogapen,zlogbv)) |>
  ggplot(aes(x=value, col=name))+
  geom_density()
```

However in these cases the relationship between the variables may be ambiguous and underexlored. In particular, we may be interested in how these two variables relate to each other. This can be achieved by visualizing the joint distribution with a 2d density plot. This can be done in multiple ways, for instance, we can use hexagons to count the number of occurrences within each joint area.

```{r}
trans_mw |> 
  ggplot(aes(x=zlogbv, y=zlogapen))+ 
  geom_hex() -> mw_hex
mw_hex
```

In this plot we can see how BV and AE relate to each other. This is informative because the assumption with these two variables is that when participants are focused on the task, they should generate high AE score and have consistent BV. If this is the case then we would expect that BV remains around zero, and that AE is high when participants are focused on the task. On the other hand, when participants are not focused on the task, they should show more BV and lower AE. This can be gauged to some degree on the previous plot. To indicate what I mean, I have highlighted the areas corresponding to each state of mind wandering below.

```{r}
library(ggforce)
mw_hex +
  annotate("rect", xmin=1.5, xmax=6.5, ymin=-1.75,ymax=1.75, alpha=.0, col="red")+
  annotate("text", label="Mind wandering", x=4,y=1.3, alpha=1, col="red")+
  annotate("rect", xmin=-1.5, xmax=1.5, ymin=1.75,ymax=6.2, alpha=.0, col="red")+
  annotate("text", label="Task focused", x=0,y=5, alpha=1, col="red")
```

However, we do not exactly find such a distribution. Although, it is important to point out that these marked areas are quite extreme (but useful for illustrative purposes). Since the data is z-scored, we would not expect a lot of data higher than 2. Thus task focused and mind wandering should be closer, but nevertheless distributed according to the markings. At any rate, this will be explored in the forth exercise - distribution over categories (MW and/or MB). 

At any rate, although the hexagons indicate the number of observation within each value, we may benefit from using a density function instead of a count function. 

```{r}
trans_mw |> 
  filter(block=="B0") |>
  ggplot(aes(x=zlogbv, y=zlogapen)) -> joint_ae.bv
  
joint_ae.bv +
  geom_density2d()
```

With this plot we can see the probability that a data point falls within a specific value. More specifically how the joint parameter space is distributed. In line with what I wrote above, we see a slight tendency towards the distribution I outlined above. 

We can use different density functions, different binwidths or different number of bins to improve visualization.

```{r}
joint_ae.bv + 
  geom_density2d(contour_var = "count")
```

In this plot, the density function is visualized as a multiplication by the number of observations. Alternatively we could visualize with the "ndensity" which is scaled to a maximum of 1: 

```{r}
joint_ae.bv + 
  geom_density2d(contour_var = "ndensity")
```

The latter two are visualized according to changes to the density function, which may have lead to some changes in number of bins.

If we follow the basic density function, we see that most values fall within the bands of the density function. 

```{r}
joint_ae.bv + 
  geom_density2d()+
  geom_point()
```

A different way of visualizing this is to use colours for the higher densities. 

```{r}
joint_ae.bv + 
  geom_density2d_filled()
```

With this plot we can more easily see the higher probability areas over the joint parameter space. To make it clear that we are looking at the join distibution of AE and BV we can add the individual distributions to the plot by using patchworks.

```{r}
#library(patchwork)

# Joint distribution
joint_ae.bv+
  geom_density2d_filled() -> p_joint 

# BV distribution
trans_mw |> 
  ggplot(aes(zlogbv))+ 
  geom_density()+
  theme_void() -> p_bv

# AE distribution
trans_mw |> 
  ggplot(aes(zlogapen))+
  geom_density()+
  theme_void()+
  coord_flip() -> p_ae

# Combining useing patchwork
p_bv + 
  ggplot()+theme_void() + # Empty element 
  p_joint + 
  p_ae
```

Although we can see the individual distributions and the joint distribution, it is not perfect due to the way patchworks glues plots together. Another solution is to use the package "ggsides", in this library we can add a side density to the main plot with one line of code:

```{r}
# library(ggside)
joint_ae.bv + 
  geom_density_2d_filled()+ 
  geom_xsidedensity()+
  geom_ysidedensity()
```

With this library we get a more neat visualization of both the individual and joint distribution. We can now make some aesthetic changes to make it more neat. 

```{r}
min(trans_mw$zlogapen)->AE_min
min(trans_mw$zlogbv)->BV_min
joint_ae.bv+
  geom_density_2d_filled() +
  theme_minimal()+
  coord_cartesian(xlim=c(AE_min,2), ylim=c(BV_min,2.5), expand = F)+
  labs(x= "Z-scored BV", y="Z-scored AE", fill="Density")+
  geom_xsidedensity()+ 
  scale_xsidey_continuous(expand=c(0,0)) + 
  theme(ggside.panel.scale=.2)
  geom_ysidedensity() -> p_ggside
p_ggside  
```

One way to make the plot more neat is to zoom into the area that contains most of the density bandwidth. However, this cases some problems with the ggside densities, since they are also scaled to the same coordinates as the main plot. One way to solve this is to increase the range of the side densities. 

```{r}
p_ggside + 
  theme(ggside.panel.scale.y=.5,
        ggside.panel.scale.x=.5)

```

However, the problem remain. It is still hard to see the exact distributions of the densities on the side. This could plausibly be fixed by re-scaling the side plots. However, I am struggling to find the right code. 

using colours within the same graph or facets over different plots. In these cases the dimensions has to be the same or very similar. However, another way of visualizing data multivariate plots can be to use integrate two plots 

With this, I can explore some basic features of the different variables. Moreover, there are different ways of visualizing multivariate 

```{r}
trans_mw |> 
  filter(block=="B0") |>
  ggplot(aes(x=zlogbv, y=zlogapen))+ 
  geom_hex()
  geom_density()
  geom_contour()
  
trans_mw |> 
  ggplot(aes(x=logbv, y=apen))+
  geom_density_2d_filled()
  geom_point()
```

```{r}
plot_ly(trans_mw, x=zlogbv, y=zlogapen)
trans_mw |> 
  ggplot(aes(x=proberound, y=zlogbv, z=1:3509))+ 
  geom_point()
```

```{r}
trans_mw |> 
  ggplot(aes(x=proberound, y=zlogapen))+ 
  geom_point()
```




